---
title: '5\. Worksheet: Alpha Diversity'
author: 'Andrea Phillips; Z620: Quantitative Biodiversity, Indiana University'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of local or site-specific diversity, also known as alpha ($\alpha$) diversity. 
First we will quantify two of the fundamental components of ($\alpha$) diversity: **richness** and **evenness**.
From there, we will then discuss ways to integrate richness and evenness, which will include univariate metrics of diversity along with an investigation of the **species abundance distribution (SAD)**.

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) to your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Answer questions in the  worksheet.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme).
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For the assignment portion of the worksheet, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file `AlphaDiversity_Worskheet.Rmd` and the PDF output of `Knitr` (`AlphaDiversity_Worskheet.pdf`).


## 1) R SETUP

In the R code chunk below, please provide the code to: 
1) Clear your R environment,
2) Print your current working directory,
3) Set your working directory to your `5.AlphaDiversity` folder, and
4) Load the `vegan` R package (be sure to install first if you haven't already).

```{r}
rm(list=ls())
getwd()
setwd("/Users/andjr/GitHub/QB2019_Phillips/2.Worksheets/5.AlphaDiversity")
install.packages("vegan",repos = "http://cran.us.r-project.org")
require("vegan")

```

## 2) LOADING DATA

In the R code chunk below, do the following:
1) Load the BCI dataset, and 
2) Display the structure of the dataset (if the structure is long, use the `max.level = 0` argument to show the basic information).

```{r}
data(BCI)
str(BCI, max.level=0)
```

## 3) SPECIES RICHNESS

**Species richness (S)** refers to the number of species in a system or the number of species observed in a sample.

### Observed richness

In the R code chunk below, do the following:

1. Write a function called `S.obs` to calculate observed richness

2. Use your function to determine the number of species in `site1` of the BCI data set, and

3. Compare the output of your function to the output of the `specnumber()` function in `vegan`.

```{r}
S.obs <- function(x = ""){rowSums(x > 0) * 1}
site1 <- BCI[1, ]
S.obs(site1)
specnumber(site1)
site1t4 <- BCI[1:4, ]
S.obs(site1t4)
```

***Question 1***: Does `specnumber()` from `vegan` return the same value for observed richness in `site1` as our function `S.obs`?
What is the species richness of the first four sites (i.e., rows) of the BCI matrix?

> ***Answer 1***:
Yes, they both returned the same value for observed richness at site 1, of 93. 
Site 1 has a species richness of 93, Site 2 of 84, Site 3 of 90, and Site 4 of 94, for a combined species richness of 361.

### Coverage: How well did you sample your site?

In the R code chunk below, do the following:

1. Write a function to calculate Good's Coverage, and

2. Use that function to calculate coverage for all sites in the BCI matrix.

```{r}
C <- function(x = ""){1 - rowSums(x == 1) / rowSums(x)}
C(BCI)
C(site1)
```
  
***Question 2***: Answer the following questions about coverage:

a.  What is the range of values that can be generated by Good's Coverage? 
b.  What would we conclude from Good's Coverage if $n_{i}$ equaled *N*?
c.  What portion of taxa in `site1` was represented by singletons? 
d.  Make some observations about coverage at the BCI plots.

> ***Answer 2a***: Good's Coverage should yield values between 0 and 1.  

> ***Answer 2b***: If n_i equalled N in Good's Coverage, we would conclude that every species sampled was a new species.   

> ***Answer 2c***:  0.91, or 91% of the species in site 1 were singletons. 

> ***Answer 2d***:  I believe we can interpret from these Good's Coverage outputs that at all sites, between 85% and 95% of the species are singletons. This shows that the coverage was probably not great, and more effort could be put into sampling in order to give us more confidence in extrapolating to estimate species diversity. 


### Estimated richness

In the R code chunk below, do the following:

1. Load the microbial dataset (located in the `5.AlphaDiversity/data` folder),

2. Transform and transpose the data as needed (see handout),

3. Create a new vector (`soilbac1`) by indexing the bacterial OTU abundances of any site in the dataset,

4. Calculate the observed richness at that particular site, and 

5. Calculate coverage of that site

```{r}
soilbac <- read.table("data/soilbac.txt", sep = "\t", header = TRUE, row.names = 1)
soilbac.t <- as.data.frame(t(soilbac))
soilbac1 <- soilbac.t[1,]
S.obs(soilbac1)
C(soilbac1)

```

***Question 3***: Answer the following questions about the soil bacterial dataset. 

a.  How many sequences did we recover from the sample `soilbac1`, i.e. *N*? 
b.  What is the observed richness of `soilbac1`? 
c.  How does coverage compare between the BCI sample (`site1`) and the KBS sample (`soilbac1`)?

> ***Answer 3a***:  We recovered 13310 sequences, aka variables.

> ***Answer 3b***: The observed richness of soilbac1 is 1074 species. 

> ***Answer 3c***:  The coverage of soilbac1 is probably better than that of site1 of the BCI sample. soilbac1 has a coverage of around 65%, meaning that 65% of the individuals in the sample were new/singleton species, while site1 had 93% of the individuals as new/singleton species. More effort was probably put into sampling soilbac1, because fewer species were singletons.


### Richness estimators

In the R code chunk below, do the following:

1. Write a function to calculate **Chao1**,

2. Write a function to calculate **Chao2**, 

3. Write a function to calculate **ACE**, and

4. Use these functions to estimate richness at `site1` and `soilbac1`. 

```{r}
S.chao1 <- function(x = ""){S.obs(x) + (sum(x == 1)^2) / (2 * sum(x == 2))}
S.chao2 <- function(site = "", SbyS = ""){
  SbyS = as.data.frame(SbyS) 
  x = SbyS[site, ] 
  SbyS.pa <- (SbyS > 0)* 1 
  Q1 = sum(colSums(SbyS.pa) == 1) 
  Q2 = sum(colSums(SbyS.pa) == 2) 
  S.chao2 = S.obs(x) + (Q1^2)/(2 * Q2) 
  return(S.chao2)}
S.ace <- function(x = "", thresh = 10){
  x <- x[x>0]
  S.abund <- length(which(x > thresh))
  S.rare <- length(which(x <= thresh))
  singlt <- length(which(x == 1))
  N.rare <- sum(x[which(x <= thresh)])
  C.ace <- 1 - (singlt / N.rare)
  i <- c(1:thresh)
  count <- function(i, y){
    length(y[y == i])
  }
  a.1 <- sapply(i, count, x)
  f.1 <- (i * (i - 1)) * a.1
  G.ace <- (S.rare/C.ace)*(sum(f.1)/(N.rare*(N.rare-1)))
  S.ace <- S.abund + (S.rare/C.ace) + max(G.ace,0)
}
S.chao1(site1)

S.chao2(1, site1)

S.ace(site1)

S.chao1(soilbac1)

S.chao2(1, soilbac1)

S.ace(soilbac1)

```

***Question 4***: What is the difference between ACE and the Chao estimators? Do the estimators give consistent results? Which one would you choose to use and why?

> ***Answer 4***: The ACE estimator only looks at the species that have less than 10 individuals--it focuses on rare taxa. Chao looks at all the individuals, and estimates how many new species were missed by extrapolating from the number of singletons and doubletons. (Chao1 uses the observed number os singletons and doubletons, while Chao2 looks at whether species showed up once or more than once (2).) The estimators give extremely inconcsistent results, from 1 to 120 for site 1, and from 2629 to infinity for soilbac1. Which one I would use depends on the kind of data I'm looking at. If a majority of the species are rare, or showing up fewer than ten times, I would use the ACE estimator. In this case the ACE estimator outputted a 1 for site1, and an Inf for soilbac1. I would use the Chao estimators for all other data sets, so that the larger incident species are not ignored in the calculation. 



### Rarefaction

In the R code chunk below, please do the following:

1. Calculate observed richness for all samples in `soilbac`,

2. Determine the size of the smallest sample,

3. Use the `rarefy()` function to rarefy each sample to this level,

4. Plot the rarefaction results, and

5. Add the 1:1 line and label.


```{r}
soilbac.S <- S.obs(soilbac.t)
min.N <- min(rowSums(soilbac.t))
S.rarefy <- rarefy(x = soilbac.t, sample = min.N, se = TRUE)
rarecurve(x = soilbac.t, step = 20, col = "blue", cex = 0.6, las=1)
abline(0, 1, col = 'red')
text(1500, 1500, "1:1", pos = 2, col = 'red')

```


##4) SPECIES EVENNESS
Here, we consider how abundance varies among species, that is, **species evenness**.

### Visualizing evenness: the rank abundance curve (RAC)
One of the most common ways to visualize evenness is in a **rank-abundance curve** (sometime referred to as a rank-abundance distribution or Whittaker plot).
An RAC can be constructed by ranking species from the most abundant to the least abundant without respect to species labels (and hence no worries about 'ties' in abundance). 

In the R code chunk below, do the following:

1. Write a function to construct a RAC,

2. Be sure your function removes species that have zero abundances, 

3. Order the vector (RAC) from greatest (most abundant) to least (least abundant), and 

4. Return the ranked vector

```{r}
RAC <- function(x = ""){
  x = as.vector(x)
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  return(x.ab.ranked)
}
```

Now, let's examine the RAC for `site1` of the BCI data set.

In the R code chunk below, do the following:

1. Create a sequence of ranks and plot the RAC with natural-log-transformed abundances,

2. Label the x-axis "Rank in abundance" and the y-axis "log(abundance)"

```{r}
plot.new()
site1 <- BCI[1, ]

rac <- RAC(x = site1)
ranks <- as.vector(seq(1, length(rac)))
opar <- par(no.readonly = TRUE)
par(mar = c(5.1, 5.1, 4.1, 2.1))
plot(ranks, log(rac), type = 'p', axes = F, xlab = "Rank in abundance", ylab = "Abundance", las = 1, cex.lab = 1.4, cex.axis = 1.25)
box()
axis(side = 1, labels = T, cex.axis = 1.25)
axis(side = 2, las = 1, cex.axis = 1.25, labels = c(1, 2, 5, 10, 20), at = log(c(1, 2, 5, 10, 20)))

```

***Question 5***: What effect does visualizing species abundance data on a log-scaled axis have on how we interpret evenness in the RAC?

> ***Answer 5***: This shows us how unequally distributed the abundance is in the sample. There are species with varying abundances across the board, showing that this site displays great unevenness. 


Now that we have visualized unevennes, it is time to quantify it using Simpson's evenness ($E_{1/D}$) and Smith and Wilson's evenness index ($E_{var}$).

### Simpson's evenness ($E_{1/D}$)

In the R code chunk below, do the following:

1. Write the function to calculate $E_{1/D}$, and

2. Calculate $E_{1/D}$ for `site1`.

```{r}
SimpE <- function(x = ""){
  S <- S.obs(x)
  x = as.data.frame(x)
  D <- diversity(x, "inv")
  E <- (D)/S
  return(E)
}

site1 <- BCI[1, ]
SimpE(site1)
```

### Smith and Wilson's evenness index ($E_{var}$)

In the R code chunk below, please do the following:

1. Write the function to calculate $E_{var}$,

2. Calculate $E_{var}$ for `site1`, and

3. Compare $E_{1/D}$ and $E_{var}$.

```{r}
Evar <- function(x){
  x <- as.vector(x[x > 0])
  1 - (2/pi)*atan(var(log(x)))
}
Evar(site1)
```

***Question 6***: Compare estimates of evenness for `site1` of BCI using $E_{1/D}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 6***: Simpson's Index returned an evenness value of approximately 0.42. Smith and wilson's returned slightly higher evenness, of about 0.51. Therefore, Smith and Wilson's says that site1 is more even that Simpson's does. This shows that the more abundant species were probably biasing the results obtained from Simpson's, and Smith and wilson's decreased the effect of that bias to show a more even spread of species abundance. 


##5) INTEGRATING RICHNESS AND EVENNESS: DIVERSITY METRICS

So far, we have introduced two primary aspects of diversity, i.e., richness and evenness.
Here, we will use popular indices to estimate diversity, which explicitly incorporate richness and evenness
We will write our own diversity functions and compare them against the functions in `vegan`. 

### Shannon's diversity (a.k.a., Shannon's entropy)

In the R code chunk below, please do the following:

1. Provide the code for calculating H' (Shannon's diversity),

2. Compare this estimate with the output of `vegan`'s diversity function using method = "shannon".

```{r}
ShanH <- function(x = ""){
  H = 0
  for (n_i in x){
    if(n_i > 0) {
      p = n_i / sum(x)
      H = H - p*log(p)
    }
  }
  return(H)
}

diversity(site1, index = "shannon")

```

### Simpson's diversity (or dominance)

In the R code chunk below, please do the following:

1. Provide the code for calculating D (Simpson's diversity),

2. Calculate both the inverse (1/D) and 1 - D,

2. Compare this estimate with the output of `vegan's` diversity function using method = "simp".

```{r}
SimpD <- function(x = ""){
  D = 0
  N = sum(x)
  for (n_i in x){
    D = D + (n_i^2)/(N^2)
  }
  return(D)
}
D.inv <- 1/SimpD(site1)
D.sub <- 1-SimpD(site1)

diversity(site1, "inv")
diversity(site1, "simp")
```

***Question 7***: Compare estimates of evenness for `site1` of BCI using $E_{H'}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 7***: EH (Shannon's Diversity?) Gives a value of 4.0 for a diversity index, while Evar gives a value of 0.51 for evenness. These values can technically be said to be in agreeance--a fairly high diversity index can correspond to moderate evenness, as long as the spread of species discovery is maintained and no one species appears drastically more often than others. 

### Fisher's $\boldsymbol\alpha$

In the R code chunk below, please do the following:

1. Provide the code for calculating Fisher's $\boldsymbol\alpha$,

2. Calculate Fisher's $\boldsymbol\alpha$ for `site1` of BCI.

```{r}
rac <- as.vector(site1[site1 > 0])
invD <- diversity(rac, "inv")
invD

Fisher <- fisher.alpha(rac)
Fisher
```

***Question 8***: How is Fisher's $\boldsymbol\alpha$ different from $E_{H'}$ and $E_{var}$? What does Fisher's $\boldsymbol\alpha$ take into account that $E_{H'}$ and $E_{var}$ do not?

> ***Answer 8***: Fisher's attempts to estimate diversity by looking at the number of species and the number of individuals in those species, rather than calculate a diversity metric that gives a snapshot view of the whole site. 

##6) MOVING BEYOND UNIVARIATE METRICS OF $\alpha$ DIVERSITY

The diversity metrics that we just learned about attempt to integrate richness and evenness into a single, univariate metric.
Although useful, information is invariably lost in this process. 
If we go back to the rank-abundance curve, we can retrieve additional information -- and in some cases -- make inferences about the processes influencing the structure of an ecological system.

## Species abundance models

The RAC is a simple data structure that is both a vector of abundances.
It is also a row in the site-by-species matrix (minus the zeros, i.e., absences). 

Predicting the form of the RAC is the first test that any biodiversity theory must pass and there are no less than 20 models that have attempted to explain the uneven form of the RAC across ecological systems.

In the R code chunk below, please do the following:

1. Use the `radfit()` function in the `vegan` package to fit the predictions of various species abundance models to the RAC of `site1` in BCI,

2. Display the results of the `radfit()` function, and 

3. Plot the results of the `radfit()` function using the code provided in the handout.

```{r}
RACresults <- radfit(site1)
plot.new()
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25) 



```


***Question 9***: Answer the following questions about the rank abundance curves:
a) Based on the output of `radfit()` and plotting above, discuss which model best fits our rank-abundance curve for `site1`?
b) Can we make any inferences about the forces, processes, and/or mechanisms influencing the structure of our system, e.g., an ecological community?

> ***Answer 9a***:  I would say that the Mandelbrot model best fits our RAC for site1. 
> ***Answer 9b***: Considering the Mandelbrot model controls for evenness, so we might be able to say that our ecological community is influenced by processes that lead to some species becoming disproportionately more abundant than others (at least I think that's what this means). So, we may have an ecosystem that lends itself to some species specializing to the environment and becoming abundant and outcompeting other species. This may mean there are fewer niches available for other species to diversify to fill. 


***Question 10***: Answer the following questions about the preemption model:
a.  What does the preemption model assume about the relationship between total abundance (*N*) and total resources that can be preempted?
b.  Why does the niche preemption model look like a straight line in the RAD plot?

> ***Answer 10a***:  It assumes that as N increases, the total resources that can be preemted decreases. 
> ***Answer 10b***:  It looks like a straight line because it draws a linear relationship between abundance and resources. 



***Question 11***: Why is it important to account for the number of parameters a model uses when judging how well it explains a given set of data?

> ***Answer 11***: I think if too many parameters are used to explain a given set of data, you might get a little convoluted trying to figure out what is actually affecting the patterns, sort of like trying to control for too many variables. Additionally, if too few parameters are used to explain a data set, you may end up missing some things that may be coming into play. 


## SYNTHESIS

1.  As stated by Magurran (2004) the  ${D = } \sum p_i^2$ derivation of Simpson's Diversity only applies to communities of infinite size. For anything but an infinitely large community, Simpson's Diversity index is calculated as ${D = } \sum \frac{n_i(n_i -1)} {N(N-1)}$. 
Assuming a finite community, calculate Simpson's D, 1 - D, and Simpson's inverse (i.e. 1/D) for `site 1` of the BCI site-by-species matrix. 

>SimpD(site1)
[1] 0.0253707

>D.sub <- 1-SimpD(site1)
> D.sub
[1] 0.9746293

> D.inv <- 1/SimpD(site1)
> D.inv
[1] 39.41555

2.  Along with the rank-abundance curve (RAC), another way to visualize the distribution of abundance among species is with a histogram (a.k.a., frequency distribution) that shows the frequency of different abundance classes.
For example, in a given sample, there may be 10 species represented by a single individual, 8 species with two individuals, 4 species with three individuals, and so on.
In fact, the rank-abundance curve and the frequency distribution are the two most common ways to visualize the species-abundance distribution (SAD) and to test species abundance models and biodiversity theories.
To address this homework question, use the R function **hist()** to plot the frequency distribution for `site 1` of the BCI site-by-species matrix, and describe the general pattern you see.

> site1hist<-t(site1)
> site1hist<-as.data.frame(site1hist)
> hist(site1hist[,1])

>The general pattern here is that there are many, many incidences (about 175) of few individuals (0-2), and very quickly we get to where there are very few incidences where more individuals than just a few in a given species show up. What we end up with is a heavily right-skewed distribution. 

3.  We asked you to find a biodiversity dataset with your partner.
This data could be one of your own or it could be something that you obtained from the literature. 
Load that dataset.
How many sites are there?
How many species are there in the entire site-by-species matrix?
Any other interesting observations based on what you learned this week?

> install.packages("OTUtable")
> require("OTUtable")
> data(otu_table)

> This site-by-species matrix has 6208 sites, and 1387 species. 
> I ran a quick Good's Coverage of the whole dataset, and just from my cursory look it appears that the coverage of this data set is better than the BCI set we looked at--a lot of the OTU sites have lower Good's Coverage values on average than the BCI data. Just for practice I also calculated the number of species in the first row of the table, and came up with the same value for both the S.obs function and the specnumber function from the vegan package. 

## SUBMITTING YOUR ASSIGNMENT
Use Knitr to create a PDF of your completed alpha_assignment.Rmd document, push it to GitHub, and create a pull request.
Please make sure your updated repo include both the HTML and RMarkdown files.

Unless otherwise noted, this assignment is due on **Wednesday, January 23^rd^, 2017 at 12:00 PM (noon)**.